Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-28 paper/manuscript_for_final_submission/manuscript_nofigs_embeddedtabsrefs.tex
changes requested by _eLife_ editorial staff
Convert Figure supplements with tabular data
into source data.

Add references to all supplements in figure legends.

@@

\begin{figure}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in \FIG{tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in \FIG{tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in \FIGSUPP[tree]{identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in \FIG{tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in \FIG{tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

\begin{figure}
%\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{}%\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

\begin{figure}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIGSUPP[mutfreqs]{mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
\FIGSUPP[mutfreqs]{contour} shows the correlations using contour rather than scatter plots.
}
\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs}.
Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
{}
\figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
{
This figure shows the same data as in \FIG{mutfreqs}B except that it shows KDE contour plots rather than points.
As is obvious from this representation, the vast majority of the points fall near the origin (very low preference in both replicates), and most of the correlation signal is therefore due to the relatively modest number of amino acids that have high preference at any given site.
This is expected, since most amino acid mutations at most sites will be strongly deleterious, and so have low preference.
}{}%\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\end{figure}

The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
\FIG{mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (\FIG{mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIGSUPP[mutfreqs]{mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIGSUPP[mutfreqs]{mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
\FIG{mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in \FIG{dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (\FIG{mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs---something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIGSUPP[mutfreqs]{mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

\begin{figure}
\begin{fullwidth}
%{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}.
We report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of $\omega_r$ are unreliable for individual sites due to low numbers of observations, making the $P$-value a better indicator of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}.
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
%{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
\begin{fullwidth}
\caption{\label{fig:prefsdist}
Env sites with shifted amino-acid preferences between BG505 and BF520.
Note that the preferences have been re-scaled using the stringency parameters in \TABLE{phydms} to enable direct comparison across Envs.
{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
We have triplicate measurements for each Env.
We calculate the distance between each pair of replicate measurements, and group these into comparisons \emph{between} the two Envs and \emph{within} replicates for the same Env.
We compute the root-mean-square distance (RMSD) for both sets of comparisons, which we denote as RMSD$_{\rm{between}}$ and RMSD$_{\rm{within}}$.
The latter quantity is a measure of experimental noise.
The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites.
The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin (HA) protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
Using this approach, 30 of the 659 sites have significant shifts (corrected distance $\ge$0.22).
{\bf (C)} All sites that have significantly shifted their amino-acid preferences at an FDR of 0.01.
For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
The wildtype amino acid for that Env is indicated using the small black letters above each logo plot; note how the wildtype amino acid is frequently but not always the most preferred one.
The sites are sorted by the magnitude of the shift.
}
\figdata{\label{figdata:prefsdist}The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
\end{fullwidth}
\end{figure}

@@

There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIGSUPP[shiftfeatures]{subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
That fact that Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open} may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

>>>>

% do not show figures
\renewcommand\includegraphics[2][]{}

@@

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in \FIG{tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in \FIG{tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in \FIGSUPP[tree]{identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in \FIG{tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in \FIG{tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

\begin{figure}
%\begin{minipage}[t]{0.34\textwidth}
%{\bf \Large A} \\ 
%\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
%\centerline{\small BG505}
%\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
%\centerline{\small BF520}
%\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
%\end{minipage}
%\begin{minipage}[t]{0.01\textwidth}
%\end{minipage}
%\begin{minipage}[t]{0.65\textwidth}
%{\bf \Large B} \\
%\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
%\end{minipage}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
\FIGSUPP[mutfreqs]{contour} shows the correlations using contour rather than scatter plots.
}
\figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
{
This figure shows the same data as in \FIG{mutfreqs}B except that it shows KDE contour plots rather than points.
As is obvious from this representation, the vast majority of the points fall near the origin (very low preference in both replicates), and most of the correlation signal is therefore due to the relatively modest number of amino acids that have high preference at any given site.
This is expected, since most amino acid mutations at most sites will be strongly deleterious, and so have low preference.
}{\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\figdata{\label{figdata:mutfreqs} Average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs} are in \texttt{avgmutfreqs.csv}.
There is only one \emph{DNA} sample for BF520 which is listed three times with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} minus \emph{DNA}, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} minus the \emph{virus}.
We use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection.}
\end{figure}

The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
\FIG{mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (\FIG{mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
\FIG{mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in \FIG{dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B, and the preferences themselves are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (\FIG{mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs---something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}.
We report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of $\omega_r$ are unreliable for individual sites due to low numbers of observations, making the $P$-value a better indicator of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}.
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
The numerical values of the preferences are in \FIG{BG505prefs}-source~data~\ref{figdata:BG505prefs}, the mapping from sequential to HXB2 numbering is in \FIG{BG505prefs}-source~data~\ref{figdata:BG505numbering}, and the $\omega_r$ values are in \FIG{BG505prefs}-source~data~\ref{figdata:BG505omega}.
}
\figdata{\label{figdata:BG505prefs}The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{\label{figdata:BG505numbering}The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{\label{figdata:BG505omega}The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
The numerical values of the preferences are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520prefs}, the mapping from sequential to HXB2 numbering is in \FIG{BF520prefs}-source~data~\ref{figdata:BF520numbering}, and the $\omega_r$ values are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520omega}.
}
\figdata{\label{figdata:BF520prefs}The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{\label{figdata:BF520numbering}The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{\label{figdata:BF520omega}The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
\begin{fullwidth}
%{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
%\hspace{0.01\textwidth} 
%\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
%\hspace{0.03\textwidth}
%\includegraphics[clip=true, trim=0in 0in 0in 0.4in, width=0.65\textwidth]{figures/distance_distribution.pdf}
%\vspace{0.15in}
%\\ 
%{\bf \Large C} \\
%\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
\caption{\label{fig:prefsdist}
Env sites with shifted amino-acid preferences between BG505 and BF520.
Note that the preferences have been re-scaled using the stringency parameters in \TABLE{phydms} to enable direct comparison across Envs.
{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
We have triplicate measurements for each Env.
We calculate the distance between each pair of replicate measurements, and group these into comparisons \emph{between} the two Envs and \emph{within} replicates for the same Env.
We compute the root-mean-square distance (RMSD) for both sets of comparisons, which we denote as RMSD$_{\rm{between}}$ and RMSD$_{\rm{within}}$.
The latter quantity is a measure of experimental noise.
The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites (see \FIG{prefsdist}-source~data~\ref{figdata:prefsdist} for numerical values).
The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin (HA) protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
Using this approach, 30 of the 659 sites have significant shifts (corrected distance $\ge$0.22).
{\bf (C)} All sites that have significantly shifted their amino-acid preferences at an FDR of 0.01.
For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
The wildtype amino acid for that Env is indicated using the small black letters above each logo plot; note how the wildtype amino acid is frequently but not always the most preferred one.
The sites are sorted by the magnitude of the shift.
}
\figdata{\label{figdata:prefsdist}The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
\end{fullwidth}
\end{figure}

@@

There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIG{shiftfeatures}-source~data~\ref{figdata:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
That fact that Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open} may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
